---
title: "R6 Introduction"
author: "Greg Snow"
date: "September 26, 2019"
output: 
  revealjs::revealjs_presentation:
    reveal_options:
      slideNumber: false
      controls: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(R6)
library(fortunes)
options(width=70)
```

## Law of Programming {data-background="IHR_Background.jpg"}

\  

Every program can be shortened by at least one line and has at least one bug.

\  

By induction, every program can be reduced to 1 line that does not work.

# Functional vs. Object Oriented programming {data-background="IHR_Background.jpg"}

## Functional Programming {data-background="IHR_Background.jpg"}

* Function results based only on inputs.

  - Same inputs imply same output, regardless of other variables/options.
  
* Functions only return results, do not change inputs.

  - Inputs may be copied internally.
  
* Clean, simple, predictable.

  - Sometimes redundant, inefficient, and a pain to pass everything around.
  
## Pass by Reference {data-background="IHR_Background.jpg"}

* Instead of passing in data, pass the address in memory where the data is located

* Function can access data without copying

* Function can also change the data in place

* Less redundant, more efficient, less of a pain in some ways

  - Not clean, Not simple, Not predictable

## Object Oriented Programming {data-background="IHR_Background.jpg"}

* Data/Object focused, methods act on specific objects

* Object can be modified by its methods

* Can restrict changing object to methods to avoid some reference problems

## {data-background="IHR_Background.jpg"}

S/R is a mixture of all these (possibly plus others).

```{r fortune1}
fortune(322)
```

## Why use objects? {data-background="IHR_Background.jpg"}

* Keep related variables together.

* Keep information between function calls.

* Avoid global variables and options.

* All the cool kids are doing it.

# Global variables {data-background="IHR_Background.jpg"}

## Quick Quiz {data-background="IHR_Background.jpg"}

```{r qq, eval=FALSE}
x <- 10
y <- FunctionThatUsesGlobalVariables(x)
```

What is the value of `x`?

## {data-background="IHR_Background.jpg"}

```{r fortune2}
fortune(174)
```

## {data-background="IHR_Background.jpg"}

```{r fortune3}
fortune(236)
```

## {data-background="IHR_Background.jpg"}

```{r fortune4}
fortune(106)

fortune(181)
```

# Quick History {data-background="IHR_Background.jpg"}

## S3 {data-background="IHR_Background.jpg"}

Version 3 of the original S language included generic functions.

* Data object can have a "class".

* Generic functions call another function/method based on the class.

  - Quick and simple
  
  - No checking to make sure objects have the correct fields
  
## S4 {data-background="IHR_Background.jpg"}

Version 4 of S implemented a new object oriented paradigm.

* Objects and methods must now be registered

* Safer because objects guaranteed to have correct fields

* More work up front

* Less flexible

## Reference Classes {data-background="IHR_Background.jpg"}

R introduced Reference classes that are more like OOP languages like smalltalk, Java, C++, etc.

* Methods belong to an object rather than being generic

* Based on S4

* Still Somewhat complicated

Some people started calling these R5 objects, but the authors did not like that.

## R6 {data-background="IHR_Background.jpg"}

A new package that implements Reference Class objects based on S3 objects in a simple but powerful way.

# R6 objects {data-background="IHR_Background.jpg"}

## Creating an object {data-background="IHR_Background.jpg"}

* Assign the result of `R6Class` to a name.

* Pass lists of fields (variables) and methods (functions).

* Use `new` method and save result.

* Use `self$<name>` and `private$<name>` to reference fields. 

## public {data-background="IHR_Background.jpg"}

* List of fields and methods that will be visible and modifiable when using the object

## private {data-background="IHR_Background.jpg"}

* List of fields and methods that can be accessed from private and public methods, but not visible or modifiable from outside.

## active  {data-background="IHR_Background.jpg"}

* List of active bindings, functions that can be treated like variables.

## Special methods {data-background="IHR_Background.jpg"}

* initialize: run when a new object is created

* finalize: run when an object is removed

* clone: used to make a copy of the object

* print: display object summary or contents

# Progress bar example {data-background="IHR_Background.jpg"}

## Functional form {data-background="IHR_Background.jpg"}

```{r fpb, eval=FALSE}
student_sim <- function(n){
  x <- rnorm(n)
  mean(x)/(sd(x)/sqrt(n))
}

pb <- winProgressBar(max=10000)
setWinProgressBar(pb, value=0)
out <- replicate(10000, {
  i <- getWinProgressBar(pb) + 1
  setWinProgressBar(pb, value=i)
  student_sim(25)
})
close(pb)
```

## R6 progress bar wrapper {data-background="IHR_Background.jpg"}

```{r r6pb, eval=FALSE}
ProgBar <- R6Class("ProgBar",
                   public = list(
                     initialize = function(max,...) {
                       private$pb <- winProgressBar(max=max)
                     },
                     inc = function(amount=1) {
                       private$i <- private$i + amount
                       setWinProgressBar(private$pb, value=private$i)
                       invisible(self)
                     },
                     reset = function(value=0) {
                       private$i <- value
                       setWinProgressBar(private$pb, value=private$i)
                       invisible(self)
                     },
                     val=function() {
                       private$i
                     }
                   ),
                   private = list(
                     pb = NULL,
                     i = 0,
                     finalize = function(...) {
                       close(private$pb)
                     }
                   )
                )
```

## R6 Progress bar example {data-background="IHR_Background.jpg"}

```{r r6pb2, eval=FALSE}
pb <- ProgBar$new(10000)

pb$reset()
out <- replicate(10000, {
  pb$inc()
  student_sim(25)
})

rm(pb)
gc()
```

## Multiple bars for nested simulation {data-background="IHR_Background.jpg"}

```{r r6pb3, eval=FALSE}
n.vec <- c(5, 10, 15, 20, 25, 30, 40, 50, 75, 100)

pb.o <- ProgBar$new(length(n.vec))
pb.o$reset()

pb.i <- ProgBar$new(10000)

out <- lapply(n.vec, function(n) {
  pb.o$inc()
  pb.i$reset(0)
  replicate(10000, {
    pb.i$inc()
    student_sim(n)
  })
})

rm(pb.o, pb.i)
gc()
```

## Include function in object {data-background="IHR_Background.jpg"}

```{r r6pb4, eval=FALSE}
ProgBar$set('public', "StuSim", function(n){
  self$inc()
  student_sim(n)
})

pb <- ProgBar$new(10000)

pb$reset()
out <- replicate(10000, pb$StuSim(25))

rm(pb)
```

## Chaining {data-background="IHR_Background.jpg"}

```{r r6pbchain, eval=FALSE}
pb <- ProgBar$new(100)
pb$val()
# 0

pb$inc()$inc()$inc()$val()
# 3

pb$reset(50)$val()
# 50

pb$reset()$val()
# 0

rm(pb)
```

## Package purrrogress {data-background="IHR_Background.jpg"}

The purrrogress package does a lot of this for you (and more). 

# Memoizing {data-background="IHR_Background.jpg"}

## Memoizing {data-background="IHR_Background.jpg"}

```{r memo}
FibGen <- R6Class("FibGen",
                  public=list(
                    curmax=2,
                    fibvals = c(1,1),
                    fib = function(n) {
                      if(n <= self$curmax) {
                        return(self$fibvals[n])
                      } else {
                        cat("generating values", self$curmax+1, 'to', n, "\n")
                        for(i in (self$curmax+1):n) {
                          self$fibvals[i] <- self$fibvals[i-1] + self$fibvals[i-2]
                        }
                        self$curmax <- n
                        return(self$fibvals[n])
                      }
                    }
                  ))
```


## Memoizing  {data-background="IHR_Background.jpg"}

```{r memo2}
myfib <- FibGen$new()
myfib$fib(5)
myfib$fib(3)
myfib$fib(10)
myfib$fib(7)
myfib$fibvals
```

# Keeping function values {data-background="IHR_Background.jpg"}

## Optim Example  {data-background="IHR_Background.jpg"}

```{r optim1}
RBF <- R6Class('RBF',
               public = list(
                 x1 = numeric(0),
                 x2 = numeric(0),
                 fr = function(x) {
                   x1 <- x[1]
                   x2 <- x[2]
                   self$x1 <- c(self$x1, x1)
                   self$x2 <- c(self$x2, x2)
                   100 * (x2 - x1*x1)^2 + (1-x1)^2
                 }
               ))
rbf <- RBF$new()

(res <- optim(c(-1.2, 0), rbf$fr))
```

## Path  {data-background="IHR_Background.jpg"}

```{r optpath, fig.show='hide'}
tmp1 <- seq(-1.25, 1.1, length.out=100)
tmp2 <- seq(-0.1, 1.15, length.out=100)
z <- outer(tmp1, tmp2, function(x1,x2) 100*(x2-x1*x1)^2 + (1-x1)^2)
image(tmp1, tmp2, z)
points(c(-1.2,1),c(0,1), col=c('green','red'), pch=16, cex=2)
lines(rbf$x1, rbf$x2)
```

## Path  {data-background="IHR_Background.jpg"}

```{r optpath, echo=FALSE}
```

# Active Bindings {data-background="IHR_Background.jpg"}

## Mimic syntax like `x++` and `x += y` from Perl/C++/etc. {data-background="IHR_Background.jpg"}

```{r inc1}
Inc <- R6Class('Inc',
               public=list(
                 value=0
               ),
               active = list(
                 pp = function(value) {
                   if(missing(value)) { #x++ (actually ++x)
                     self$value <- self$value + 1
                     return(self$value)
                   } else {  # x += y
                     self$value <- self$value + value
                   }
                 }
               )
               )
```

## Example {data-background="IHR_Background.jpg"}

```{r inc2}
x <- Inc$new()
x$value <- 5
x$pp
x$pp <- 10
x$value
```

## Example {data-background="IHR_Background.jpg"}

```{r inc3, eval=FALSE}
i <- Inc$new()

x <- 1e7
while( abs(x-1) > 0.001 ) {
  x <- sqrt(x)
  if( i$pp > 10 ) {
    cat("Maximum Iterations Reached\n")
    break
  }
}

for(k in longvector) {
  setWinProgressBar(pb, value=i$pp)
  LongRunningFunction(k)
}
```





# Adaptive Integration example (recursion through looping) {data-background="IHR_Background.jpg"}

## Trapazoid object {data-background="IHR_Background.jpg"}

```{r trapobject}
TrapObj <- R6Class(
  "TrapObj",
  public = list(
    x.left = numeric(1),
    y.left = numeric(1),
    x.right = numeric(1),
    y.right = numeric(1),
    split = function(fun){
      left <- self$clone()
      right <- self$clone()
      new.x <- (self$x.left + self$x.right)/2
      new.y <- if(missing(fun)) {
        (self$y.left + self$y.right)/2
      } else {
        fun(new.x)
      }
      left$x.right <- right$x.left <- new.x
      left$y.right <- right$y.left <- new.y
      list(left=left, right=right)
    },
    plot = function(col=NA){
      polygon(c(self$x.left, self$x.left, self$x.right, self$x.right),
              c(0,self$y.left, self$y.right, 0), border='blue', col=col)
    }
  ),
  active = list(
    area = function() (self$x.right - self$x.left) * 
      (self$y.right + self$y.left)/2
  )
)
```

## Stack {data-background="IHR_Background.jpg"}

```{r stackobject}
Stack <- R6Class(
  "Stack",
  public = list(
    length = 0L,
    stack = vector(mode='list', length=10),
    push = function(obj) {
      if( length(self$stack) <= self$length ) {
        self$stack[2*self$length] <- list(NULL)
      }
      self$length <- self$length + 1
      self$stack[[self$length]] <- obj
      invisible(self)
    },
    pop = function() {
      if(self$length < 1) stop('Stack is empty')
      tmp <- self$stack[[self$length]]
      self$length <- self$length - 1
      return(tmp)
    }
  )
)
```


## Adaptive Integration {data-background="IHR_Background.jpg"}

```{r adaptive, fig.show='hide'}
cand <- Stack$new()
res <- Stack$new()

#myfun <- function(x) dnorm(x, 0, 1)
myfun <- function(x) sin(x^2) + 1.25

start <- TrapObj$new()
start$x.left <- 0
start$y.left <- myfun(0)
start$x.right <- 2.5
start$y.right <- myfun(2.5)

cand$push(start)
area <- 0
tol <- 0.0005

while(cand$length){
  curve(myfun(x), from=0, to=2.5, col='lightgrey', lwd=5, ylim=c(0,2.5))
  for(i in seq_len(cand$length)) cand$stack[[i]]$plot("#ff000033")
  for(i in seq_len(res$length)) res$stack[[i]]$plot("#ff000033")
  cur <- cand$pop()

  tmp <- cur$split(myfun)
  for(i in 1:2) tmp[[i]]$plot("#0000ff33")
  
  if(abs(cur$area - tmp$left$area - tmp$right$area) <= tol) {
    res$push(tmp$left)$push(tmp$right)
    area <- area + tmp$left$area + tmp$right$area
  } else {
    cand$push(tmp$right)$push(tmp$left)
  }
}
```

## Adaptive Integration {data-background="IHR_Background.jpg"}

```{r adaptive, echo=FALSE, fig.show='animate', animation.hook=knitr::hook_gifski, interval=0.4}
```

## Results {data-background="IHR_Background.jpg"}

```{r intres, results='hide', fig.show='hide'}
curve(myfun(x), from=0, to=2.5, col='green', lwd=5, ylim=c(0,2.5))
for(i in seq_len(res$length)) {
  res$stack[[i]]$plot()
}
text(0.5, 2.3, paste('Est:',round(area,4)), adj=1)
text(0.5, 2.1, paste('Int:',round(integrate(myfun, lower=0, upper=2.5)$value,4)), adj=1)
```

## Results {data-background="IHR_Background.jpg"}

```{r intres, echo=FALSE}
```

## R6DS {data-background="IHR_Background.jpg"}

The R6DS (Data Structures) package implements stacks, queues, and other data structures using R6 objects, probably better than I have here.

# Not covered {data-background="IHR_Background.jpg"}

## Not covered {data-background="IHR_Background.jpg"}

* Inheritance

* Debugging

* Deep Cloning (reference objects)

* Non-portable R6 objects

* Lots of theory and other things


# Questions {data-background="IHR_Background.jpg"}

